 

### 2021-05-05 vertical-thermal-review.Rmd
small edits

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al. 2014*].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively relative to pine and boreal forests, and compared to nearby clearings. These forests maintain cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively. [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. Similarly, when compared to neighboring tree crop agricultural plantations, intact tropical forests maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity with greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009]. Therefore, they maintain higher daytime maximum temperature and light availability in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009]. During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil from daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Buffering is also affected by Canopy roughness, turbulent air flow and the canopy boundary layer. As a result, taller trees upto 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaud [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019]. Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]

@@

**Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
For leaves, maintaining optimum leaf temperature for metabolic optima in relation to the biophysical environment, maximizes carbon assimilation in a given circumstance. This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`). The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015]. When leaves are cooler than air at high temperature and warmer than air at low temperature, they considered to exhibit limited homeothermy. When leaves are drastically warmer than $T_{air}$ such as in the upper canopy, this is considered megathermy. In some cases, $T_{leaf}$ can be equalled to $T_{air}$, showing poikiolothermic behavior [Cavaleri, 2020].

@@

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation and water availability, evaporative cooling sustains $T_{leaf}$ under increased $T_{air}$ [@cavaleriColdbloodedForestsWarming2020] until stomtal limitation impedes cooling, which causes leaf heat stress [@fauset_differences_2018].

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al. 2018].  However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low. This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018]. On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
Canopy buffering capacity, as well vertical biophysical gradient, affects vertical distribution of leaf temperature. While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.

Particularly, in open forests with sparse distribution of trees and lower buffering capacity, as observed in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. Greater heat stress and $T_{air}$ in the lower canopy is also observed in open Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987], this is analogous to greenhouse effect caused under dense lower canopies.  Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016]. In other tropical dry conditions, inner canopy can be warmer than above canopy and air temperature due to conditions of greater relative humidity [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

@@

@@

- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues. I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."

@@

An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [Chen et al. 2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves. Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016]. However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)

@@

Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air. [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al. 1999; @niinemetsAdjustmentFoliageStructure1998]. Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al. 2010]. On the other hand, leaf temperature increases with effective leaf width and leaf size [Fauset et al. 2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017].

@@

Ontogenetic differences also suggests shade tolerators have lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities than in intolerators [@niinemetsControversyTraitsConferring2006]. While shade intolerant species experience greater heterogeneity in light during ontogeny [poorter et al. 2013]. However, these findings conflict with  Kitajima 1994 who found for tropical seedlings; shade intolerance species grew faster in the sun and in the shade, but at the cost of higher mortality in the shade. Lawren and Karoru had contrasting findinds, and so far appears that there is no a consensus? How will leaf temperature vary between the two?

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. During the day, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]. Similarly, max transpiration rate increases with canopy height [@kosugiSeasonalFluctuationsTemperature2006]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].

@@

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)

Leaf temperature beyond photosynthetic optimum temperature, declines photosynthesis [**REF**]. Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

@@

While dark and light respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al. 2014*].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davisMicroclimaticBufferingForests2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively where they maintain cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relatively more than pine and boreal forests, and nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. Similarly, when compared to neighboring tree crop agricultural plantations, intact tropical forests maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity with greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009]. Therefore, they maintain higher daytime maximum temperature and light availability in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009]. During the night, pine forests can maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].

Buffering is also affected by canopy roughness, turbulent air flow and the canopy boundary layer. As a result, taller trees upto 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaud [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019]. Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

@@

**Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
Leaves adapt to their biophysical environment. Maintaining optimum leaf temperature for metabolic optima in relation to their biophysical environment maximizes carbon assimilation in a given circumstance. This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`). The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015]. When leaves are cooler than air at high temperature and warmer than air at low temperature, they considered to exhibit limited homeothermy. When leaves are drastically warmer than $T_{air}$ such as sometimes in the upper canopy, this is considered megathermy. In some cases, $T_{leaf}$ can be equalled to $T_{air}$, showing poikiolothermic behavior [Cavaleri, 2020; @drakeNoEvidenceHomeostatic2020].

@@

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation and water availability, evaporative cooling sustains $T_{leaf}$ under increased $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al. 2018].  However, during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low. This is caused by increase in air temperature and VPD driving greater leaf water demand than water transport, which limits stomatal conductance along with decreased leaf water content, therefore limiting heat dissapation [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018]. On the other hand, buffered leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
Canopy buffering capacity, as well vertical biophysical gradient, affect vertical distribution of leaf temperature. While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.

Particularly, in open forests with sparse distribution of trees and lower buffering capacity, as observed in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. Greater heat stress and $T_{air}$ in the lower canopy is also observed in open Australian desert trees [@curtis_intracanopy_2019] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987], this is analogous to greenhouse effect caused under dense lower canopies.  Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016]. In other tropical dry conditions, inner canopy can be warmer than above canopy and air temperature due to conditions of greater relative humidity [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

@@

**Stability of $T_{leaf}$ varies with fluctuation in biophysical conditions**
Biophysical environment fluctuation dictates how stable $T_{leaf}$ is, where stable biophysical conditions, for eg. low wind speed and consistent $T_{air}$ can prolong $T_{leaf}$, consistently reducing stomatal conductance and photosynthesis while increasing $T_{leaf}$, and therefore reducing chances for leaf recovery. However, increase in biophysical fluctuations make $T_{leaf}$ unstable as seen in upper canopies where with available water unstable $T_{leaf}$ can be more manageable to peak irradiance and $T_{air}$ than in stable buffered canopy regions [**REF**]

@@

@@

Additionally, given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016], such as by incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient, partly because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020].

@@

Along the vertical gradient, exposed leaves are better equipped to dissapate heat than shaded leaves, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air. [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Mechanisms such vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al. 1999; @niinemetsAdjustmentFoliageStructure1998]. Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al. 2010]. On the other hand, leaf temperature increases with effective leaf width and leaf size [@fauset_differences_2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]. Additionally, understory leaves tend to open their stomata wider (stomatal apeture) than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].

@@

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. During the day, max transpiration rate and stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shaded leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].

@@

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).

Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis [**REF**]. Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

@@

While respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.
